Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-07-16 paper/analysis.tex
rewrite proof slightly and add missing paragraph in results

\color{black}
\begin{proof}
    The proof proceeds by bounding moments for fixed values of $t$ and then uses Minkowski's integral inequality. We start with a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}) for which we define
    \begin{equation}
        \widetilde{\mtx{B}}(t)
        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
            \mtx{B}(t) & & \\
            & \ddots & \\
            & & \mtx{B}(t)
        \end{pmatrix}
        \quad \text{and} \quad
        \widetilde{\vct{\psi}} = \begin{pmatrix}
            \vct{\psi}_1 \\
            \vdots \\
            \vct{\psi}_{n_{\mtx{\Psi}}}
        \end{pmatrix}.
    \end{equation}
    Note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) =  \Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}}$, $\lVert \widetilde{\mtx{B}}(t) \rVert _2 = \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}}$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold. Therefore, by \reflem{lem:hutchinson-sub-exponential}, $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a centered, $(2 \lVert \mtx{B}(t) \rVert _F^2 \frac{1}{1 - 1/c} / n_{\mtx{\Psi}}, 2 \lVert \mtx{A} \rVert _2 c / n_{\mtx{\Psi}})$-sub-exponential random variable for any $c > 1$. Thus, we can apply \reflem{lem:sub-exponential-moments} to bound
    \begin{equation}
        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 2 e^{1/e} \left(\frac{\sqrt{\pi}}{1-1/c} \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 4 c \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right).
    \end{equation}
    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and letting $c = 1 + \sqrt{\pi} / 2$ we get
    \begin{equation}
        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq \underbrace{8 e^{1/e} (1 + \sqrt{\pi} / 2)}_{=~C} \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}} \right) \lVert \mtx{B}(t) \rVert _F.
        \label{equ:moment-bound-hutchinson}
    \end{equation}

>>>>

\color{black}
\begin{proof}
    The proof proceeds by bounding moments for fixed values of $t$ and then uses Minkowski's integral inequality. We start with a standard embedding trick (see, e.g.,~\cite[Theorem 1]{cortinovis-2022-randomized-trace}) for which we define
    \begin{equation}
        \widetilde{\mtx{B}}(t)
        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
            \mtx{B}(t) & & \\
            & \ddots & \\
            & & \mtx{B}(t)
        \end{pmatrix}
        \quad \text{and} \quad
        \widetilde{\vct{\psi}} = \begin{pmatrix}
            \vct{\psi}_1 \\
            \vdots \\
            \vct{\psi}_{n_{\mtx{\Psi}}}
        \end{pmatrix}.
    \end{equation}
    Note that $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) =  \Trace(\widetilde{\mtx{B}}(t)) - \widetilde{\vct{\psi}}^{\top} \widetilde{\mtx{B}}(t) \widetilde{\vct{\psi}}$, $\lVert \widetilde{\mtx{B}}(t) \rVert _2 = \lVert \mtx{B}(t) \rVert _2 / n_{\mtx{\Psi}}$ and $\lVert \widetilde{\mtx{B}}(t) \rVert _F = \lVert \mtx{B}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ hold. Therefore, by \reflem{lem:hutchinson-sub-exponential}, $\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))$ is a centered and $(2 \lVert \mtx{B}(t) \rVert _F^2 \frac{1}{1 - 1/c} / n_{\mtx{\Psi}}, 2 \lVert \mtx{A} \rVert _2 c / n_{\mtx{\Psi}})$-sub-exponential random variable for any $c > 1$. Thus, we can apply \reflem{lem:sub-exponential-moments} to bound
    \begin{equation}
        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq 2 e^{1/e} \left(\frac{\sqrt{\pi}}{1-1/c} \sqrt{\frac{p}{n_{\mtx{\Psi}}}} \lVert \mtx{B}(t) \rVert _F + 4 c \frac{p}{n_{\mtx{\Psi}}} \lVert \mtx{B}(t) \rVert _2 \right).
    \end{equation}
    Bounding $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and letting $c = 1 + \sqrt{\pi} / 2$ we get
    \begin{equation}
        \mathbb{E}^{p}[\Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t))] \leq \underbrace{8 e^{1/e} (1 + \sqrt{\pi} / 2)}_{=~C} \left( \sqrt{\frac{p}{n_{\mtx{\Psi}}}} + \frac{p}{n_{\mtx{\Psi}}} \right) \lVert \mtx{B}(t) \rVert _F.
        \label{equ:moment-bound-hutchinson}
    \end{equation}

### 2024-07-16 paper/numresults.tex
clean up bibliography and add acknowledgments
rewrite proof slightly and add missing paragraph in results

\section{Numerical results}
\label{sec:results}

@@

@@

\todo{Interpret results.}

@@

>>>>

\section{Numerical experiments}
\label{sec:results}

\subsection{Spectral density for Hamiltonian of electronic structure}
\label{subsec:hamiltonian}

@@

For a fixed budget of random vectors $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}$, \reffig{fig:distribution} visualizes how for different values of the smoothing parameter $\sigma$ the random vectors should either be invested into the Nyström approximation $n_{\mtx{\Omega}}$ or the estimation of the residual trace $n_{\mtx{\Psi}}$. Again, since for small $\sigma$ the matrix function $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ has a small numerical rank, a Nyström approximation alone is already enough to achieve a good approximation. On the other hand, when $\sigma$ is large, the Nyström approximation alone is not effective, and the correction with the residual trace becomes indispensable.

@@

\subsection{Spectral density of Hessian matrix of neural network}
\label{subsec:hessian}

@@

\subsection{Trace of parameter-dependent matrix functions}
\label{subsec:function-trace}
